Supplementary materials: $R^*$ convergence diagnostic

### 2020-10-23 src/s_cauchy.Rmd
added cauchy and sorted trends section

r_star_alt_gbm <- r_star(x, method="gbm", uncertainty = T)
r_star_alt_rf <- r_star(x, method="rf", uncertainty = T)

@@

r_star_alt_gbm <- r_star(full_data, method="gbm", uncertainty = T)
r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)
mon_alt <- monitor_extra(split_data(full_data))
```

Plot
```{r}
b_df <- tibble(nominal_gbm=r_star_nom_gbm,
               nominal_rf=r_star_nom_rf,
               alt_gbm=r_star_alt_gbm,
               alt_rf=r_star_alt_rf,
               source="r_star") %>% 
  bind_rows(tibble(nominal=mon_nom$zfsRhat[1:50],
                alt=mon_alt$zfsRhat[101:150],
                source="r_hat")) %>% 
  mutate(iter="10,000")

c_df <- a_df %>% mutate(iter="1000") %>% 
  bind_rows(b_df) %>% 
  mutate(source=if_else(source=="r_hat", "split-Rhat", source)) %>% 
  mutate(source=if_else(source=="r_star", "R*", source))
saveRDS(c_df, "cauchy_rhat_rstar.rds")
c_df <- readRDS("cauchy_rhat_rstar.rds")
my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.03) else c(0.9, 3)}

g <- c_df %>% 
  melt(id.vars=c("source", "iter")) %>% 
  mutate(source=if_else(source=="split-Rhat", "Split-Rhat", as.character(source))) %>% 
  mutate(iter=fct_relevel(as.factor(iter), "1000", "10,000")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(bins=50, position="identity", alpha=0.8) +
  scale_fill_grey("Parameterisation", labels=c("Nominal", "Alternative")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  facet_wrap(~iter+source, scales="free") +
  scale_x_continuous(limits=my_lims) +
  theme(strip.text = element_text(size=16, colour="black"),
        axis.text = element_text(size=16, colour="black"),
        axis.title = element_text(size=20, colour="black"),
        legend.text = element_text(size=16, colour="black"),
        legend.title = element_text(size=16, colour="black"),
        legend.position = "top") +
  xlab("Value")

>>>>

r_star_alt_gbm <- r_star(full_data, method="gbm", uncertainty = T)
r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)

@@

r_star_alt_gbm <- r_star(full_data, method="gbm", uncertainty = T)
r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)
mon_alt <- monitor_extra(split_data(full_data))

b_df <- tibble(nominal_gbm=r_star_nom_gbm,
               nominal_rf=r_star_nom_rf,
               alt_gbm=r_star_alt_gbm,
               alt_rf=r_star_alt_rf,
               source="r_star") %>% 
  bind_rows(tibble(nominal=mon_nom$zfsRhat[1:50],
                alt=mon_alt$zfsRhat[101:150],
                source="r_hat")) %>% 
  mutate(iter="10,000")

c_df <- a_df %>% mutate(iter="1000") %>% 
  bind_rows(b_df) %>% 
  mutate(source=if_else(source=="r_hat", "split-Rhat", source)) %>% 
  mutate(source=if_else(source=="r_star", "R*", source))
saveRDS(c_df, "cauchy_rhat_rstar.rds")
```

Plot
```{r}
c_df <- readRDS("cauchy_rhat_rstar.rds")
c_df_rhat <- c_df %>% 
  filter(source=="split-Rhat") %>% 
  select(source, nominal, alt, iter)
c_df_rstar <- c_df %>% 
  filter(source!="split-Rhat")
c_df_gbm <- c_df_rstar %>%
  select(contains("gbm"), source, iter) %>% 
  rename(nominal=nominal_gbm,
         alt=alt_gbm) %>% 
  mutate(source="R*: GBM")
c_df_rf <- c_df_rstar %>%
  select(contains("rf"), source, iter) %>% 
  rename(nominal=nominal_rf,
         alt=alt_rf) %>% 
  mutate(source="R*: RF")

my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.06) else c(0.9, 3)}
g <- c_df_rhat %>% 
  bind_rows(c_df_gbm) %>% 
  bind_rows(c_df_rf) %>% 
  melt(id.vars=c("source", "iter")) %>% 
  mutate(source=if_else(source=="split-Rhat", "Split-Rhat", as.character(source))) %>% 
  mutate(iter=fct_relevel(as.factor(iter), "1000", "10,000")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(bins=50, position="identity", alpha=0.8) +
  scale_fill_grey("Parameterisation", labels=c("Nominal", "Alternative")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  facet_wrap(~iter+source, scales="free") +
  scale_x_continuous(limits=my_lims) +
  theme(strip.text = element_text(size=16, colour="black"),
        axis.text = element_text(size=16, colour="black"),
        axis.title = element_text(size=20, colour="black"),
        legend.text = element_text(size=16, colour="black"),
        legend.title = element_text(size=16, colour="black"),
        legend.position = "top") +
  xlab("Value")

### 2020-10-23 src/s_trends.Rmd
started to change trends text
added cauchy and sorted trends section

```{r}
g <- readRDS("../output/experiment_trends_joint_distribution.rds") %>%
  as.data.frame() %>%
  select(iters, rho, rep, zsRhat, Rstar_split, method) %>%
  unique() %>%
  select(-rep) %>%
  melt(c("rho", "iters", "method")) %>%
  mutate(value=as.numeric(as.character(value))) %>% 
  mutate(variable=case_when(variable=="zsRhat"~"A. Split-Rhat",
                            variable=="Rstar_split"~"B. Split-R*")) %>% 
  mutate(iters=as.numeric(as.character(iters))) %>% 
  mutate(method=as.character(method)) %>% 
  mutate(method=if_else(method=="rf", "RF", "GBM")) %>% 
  ggplot(aes(y = value, x = rho)) +
  geom_jitter(aes(colour=as.factor(method),
                  shape=as.factor(method))) +
  facet_grid(vars(iters), vars(variable)) +
  geom_hline(yintercept = 1) +
  theme(strip.text.x = element_text(size=16, colour="black"),
        strip.text.y = element_text(size=16, colour="black"),
        axis.text.x = element_text(size=16, colour="black"),
        axis.text.y = element_text(size=16, colour="black"),
        axis.title = element_text(size=16, colour="black")) +
  ylab("Value") +
  xlab(TeX("$\\rho$"))  +
  scale_colour_grey("Classifier") +
  scale_shape("Classifier")
g

@@

>>>>

```{r}
# remove double iterates for rhat from rfs
g <- readRDS("../output/experiment_trends_joint_distribution.rds") %>%
  as.data.frame() %>%
  select(iters, rho, rep, zsRhat, Rstar_split, method) %>%
  unique() %>%
  select(-rep) %>%
  melt(c("rho", "iters", "method")) %>%
  filter(!((method=="rf")&(variable=="zsRhat"))) %>% 
  mutate(value=as.numeric(as.character(value))) %>% 
  mutate(variable=case_when(variable=="zsRhat"~"A. Split-Rhat",
                            variable=="Rstar_split"~"B. Split-R*")) %>% 
  mutate(iters=as.numeric(as.character(iters))) %>% 
  mutate(method=as.character(method)) %>% 
  mutate(method=if_else(method=="rf", "RF", "GBM")) %>% 
  ggplot(aes(y = value, x = rho)) +
  geom_jitter(aes(colour=as.factor(method),
                  shape=as.factor(method))) +
  facet_grid(vars(iters), vars(variable)) +
  geom_hline(yintercept = 1) +
  theme(strip.text.x = element_text(size=16, colour="black"),
        strip.text.y = element_text(size=16, colour="black"),
        axis.text.x = element_text(size=16, colour="black"),
        axis.text.y = element_text(size=16, colour="black"),
        axis.title = element_text(size=16, colour="black")) +
  ylab("Value") +
  xlab(TeX("$\\rho$"))  +
  scale_colour_grey("Classifier") +
  scale_shape("Classifier")
g

@@

res <- readRDS("../output/experiment_trends_rho_nearer1.rds") %>%
  mutate(method=as.character(method)) %>% 
  mutate(method=if_else(method=="gbm", "A. GBM", "B. RF"))

### 2020-10-23 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
added cauchy and sorted trends section

\begin{table}[]
	\begin{tabular}{l|l|l}
		\textbf{Example }&  \textbf{Relevance} & \textbf{Section} \\
		\midrule
		Autoregressive & \textbf{Examining $R^*$ and sensitivities to its calculation} & \ref{sec:heterogeneity} \\
		& Detecting heterogeneous chain variance using $R^*$ & \ref{sec:heterogeneity_performance} \\
		& Stochasticity in $R^*$ & \ref{sec:heterogeneity_stochasticity}\\
		& Generating $R^*$ uncertainty measure & \ref{sec:heterogeneity_uncertainty}\\
		& Sensitivity of $R^*$ to number of chains & \ref{sec:heterogeneity_numchains}\\
		& Robustness of $R^*$ to ML model hyperparameters & \ref{sec:heterogeneity_robustness}\\
		\midrule
		Multivariate normals & \textbf{Detecting convergence in joint distributions} &  \ref{sec:multivariate_normal}\\
		& Unconverged joint distribution in bivariate normal & \ref{sec:multivariate_normal_bivariate}\\
		& High correlations between dims in 250D normal  & \ref{sec:multivariate_normal_250}\\
		& Measuring contributions of variables to poor convergence & \ref{sec:multivariate_normal_varimportance}\\
		\midrule
		Cauchy & \textbf{Detecting convergence for long-tailed distributions} & \ref{sec:cauchy}\\
		& Comparing $R^*$ and existing measures to objective convergence & \ref{sec:cauchy_objective}\\
		\midrule
		Eight schools model & \textbf{Hierarchical Bayesian model slow convergence} & \ref{sec:eight_shools}\\
		\midrule 
		Wide multivariate normal & \textbf{Detecting convergence when \# draws $\sim$ \# dims} & \ref{sec:wide}\\
		\midrule
		Non-stationary marginals & \textbf{Detecting time-varying sampling distributions} & \ref{sec:non-stationary}\\
		& Trends in mean across all chains & \ref{sec:non-stationary_chains}\\
		& Trends in mean in a single dimension & \ref{sec:non-stationary_single}\\
		& Trends in covariance & \ref{sec:non-stationary_covariance}\\
		& Sensitivity of $R^*$ to chain persistence & \ref{sec:non-stationary_persistence}\\
		\midrule
		Ovarian and prostate models & \textbf{Bayesian models with many parameters and multimodal posteriors} & \ref{sec:prostate}\\
	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
\end{table}

@@

\color{red}
\subsection{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them. In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier. In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}. In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution. In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains. Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}. This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw. Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.

\subsubsection{Joint distribution}\label{sec:joint_distribution}
Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32. For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix. To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals. To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$. For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers. In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.

The results of these experiments are shown in Fig. \ref{fig:gbm_vs_rf_normal}. In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic. When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$. In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence. As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows. $R^*$ as calculated by GBMs and RFs also increases. For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification. For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares). The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians. For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process. The dashed line shows $R^*=1$. Here, $R^*$ was calculated using chains split into two halves.}
	\label{fig:gbm_vs_rf_normal}
\end{figure}

\subsubsection{Fat tails}\label{sec:tail_fatness}
We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution. To do so, we again consider four chains. In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom. In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$. We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$. We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness. Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$. To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$. For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers. Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.

The results of these experiments are shown in Fig. \ref{fig:gbm_vs_rf_studdentt}. Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions. Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$. The point colour and shadings give the method used to calculate $R^*$. In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method. As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases. In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers. In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate. If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.

Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains. The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy. In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness. These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones. This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time. Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time. RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}. Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution. The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares). The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians. For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process. The dashed line shows $R^*=1$. Here, $R^*$ was calculated using chains split into two halves.}
	\label{fig:gbm_vs_rf_studdentt}
\end{figure}

@@

In the top-left panel of Fig. \ref{fig:cauchy}, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet its mean remains above this value. In the top-right panel of Fig. \ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs. The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}<1.01$ for all parameters.

Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of 4 chains. In the bottom row of Fig. \ref{fig:cauchy}, we show the results for these longer runs. In these, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1. The $\widehat{R}$ values are all below 1.01, indicating convergence in both cases.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains. Columns show the $R^*$ distributions (left) and rank-normalised split-$\widehat{R}$ values across all parameters (right). Shadings indicate different model paramerisations as indicated in legend.}
	\label{fig:cauchy}
\end{figure}

>>>>

\begin{table}[]
	\begin{tabular}{l|l|l}
		\textbf{Example }&  \textbf{Relevance} & \textbf{Section} \\
		\midrule
		Autoregressive & \textbf{Examining $R^*$ and sensitivities to its calculation} & \ref{sec:heterogeneity} \\
		& Detecting heterogeneous chain variance using $R^*$ & \ref{sec:heterogeneity_performance} \\
		& Stochasticity in $R^*$ & \ref{sec:heterogeneity_stochasticity}\\
		& Generating $R^*$ uncertainty measure & \ref{sec:heterogeneity_uncertainty}\\
		& Sensitivity of $R^*$ to number of chains & \ref{sec:heterogeneity_numchains}\\
		\midrule
		Multivariate normals & \textbf{Detecting convergence in joint distributions} &  \ref{sec:multivariate_normal}\\
		& Unconverged joint distribution in bivariate normal & \ref{sec:multivariate_normal_bivariate}\\
		& High correlations between dims in 250D normal  & \ref{sec:multivariate_normal_250}\\
		& Measuring contributions of variables to poor convergence & \ref{sec:multivariate_normal_varimportance}\\
		\midrule
		Cauchy & \textbf{Detecting convergence for long-tailed distributions} & \ref{sec:cauchy}\\
		& Comparing $R^*$ and existing measures to objective convergence & \ref{sec:cauchy_objective}\\
		\midrule
		Eight schools model & \textbf{Hierarchical Bayesian model slow convergence} & \ref{sec:eight_shools}\\
		\midrule 
		Wide multivariate normal & \textbf{Detecting convergence when \# draws $\sim$ \# dims} & \ref{sec:wide}\\
		\midrule
		Non-stationary marginals & \textbf{Detecting time-varying sampling distributions} & \ref{sec:non-stationary}\\
		& Trends in mean across all chains & \ref{sec:non-stationary_chains}\\
		& Trends in mean in a single dimension & \ref{sec:non-stationary_single}\\
		& Trends in covariance & \ref{sec:non-stationary_covariance}\\
		& Sensitivity of $R^*$ to chain persistence & \ref{sec:non-stationary_persistence}\\
		\midrule
		Ovarian and prostate models & \textbf{Bayesian models with many parameters and multimodal posteriors} & \ref{sec:prostate}\\
		\midrule		
		\textcolor{red}{Discrete Markov model} & \textbf{Evaluating $R^*$ on discrete parameter models} & \ref{sec:discrete}\\
		& Small state-space & \ref{sec:discrete_small}\\
		& Large state-space & \ref{sec:discrete_large}\\
		\midrule
		\textcolor{red}{Various} & \textbf{Sensitivity of $R^*$ to ML model} & \ref{sec:ML_sensitivity}\\
		& Comparing different ML classifiers & \ref{sec:ml_model}\\
		& Sensitivity of $R^*$ to GBM and RF hyperparameters & \ref{sec:hyperparameters}\\
		\midrule
		\textcolor{red}{Multivariate normal \& Student-t dists.} & \textbf{Comparing GBM and RF classifiers} & \ref{sec:comparison_gbm_rf}\\
		& Detecting convergence in joint distributions & \ref{sec:joint_distribution}\\
		& Tail convergence & \ref{sec:tail_fatness}\\
	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
\end{table}

@@

@@

\textcolor{red}{In the top-left and top-middle panel of Fig. \ref{fig:cauchy}, we show the $R^*$ distribution for GBM and RF classifiers under both parameterisations.} As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet its mean remains above this value. In the top-right panel of Fig. \ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs. The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}<1.01$ for all parameters.

Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of 4 chains. In the bottom row of Fig. \ref{fig:cauchy}, we show the results for these longer runs. \textcolor{red}{In these, the alternative parameterisation now has an $R^*$ distribution centred on $R^*=1$ for the GBM classifier although the RF classifier $R^*$ distribution remains slightly rightwards of this target indicating that convergence is nearer but more iterations are likely still required.} Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1. The $\widehat{R}$ values are all below 1.01, indicating convergence in both cases.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains. Columns show the $R^*$ distributions from GBM (left) and RF (middle) and rank-normalised split-$\widehat{R}$ values across all parameters (right). Shadings indicate different model paramerisations as indicated in legend.}
	\label{fig:cauchy}
\end{figure}

### 2020-10-23 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
added cauchy and sorted trends section

>>>>

\color{red}
\section{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them. In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier. In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}. In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution. In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains. Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}. This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw. Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.

\subsection{Joint distribution}\label{sec:joint_distribution}
Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32. For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix. To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals. To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$. For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers. In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.

The results of these experiments are shown in Fig. \ref{fig:gbm_vs_rf_normal}. In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic. When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$. In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence. As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows. $R^*$ as calculated by GBMs and RFs also increases. For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification. For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares). The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians. For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process. The dashed line shows $R^*=1$. Here, $R^*$ was calculated using chains split into two halves.}
	\label{fig:gbm_vs_rf_normal}
\end{figure}

\subsection{Fat tails}\label{sec:tail_fatness}
We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution. To do so, we again consider four chains. In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom. In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$. We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$. We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness. Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$. To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$. For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers. Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.

The results of these experiments are shown in Fig. \ref{fig:gbm_vs_rf_studdentt}. Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions. Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$. The point colour and shadings give the method used to calculate $R^*$. In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method. As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases. In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers. In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate. If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.

Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains. The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy. In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness. These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones. This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time. Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time. RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}. Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution. The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares). The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians. For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process. The dashed line shows $R^*=1$. Here, $R^*$ was calculated using chains split into two halves.}
	\label{fig:gbm_vs_rf_studdentt}
\end{figure}
\color{black}
